*simex*: a disease modelling tool for decision makers
---------------------------------------------------------------

[![MIT license](https://img.shields.io/badge/License-MIT-blue.svg)](LICENSE.md) [![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://lifecycle.r-lib.org/articles/stages.html#maturing)

*simex* is an R package for simulating the spread of infectious diseases, as well as interventions such as social distancing measures, isolation and vaccination. It uses an age-structured SEIR compartmental model with country-specific age demographics and contact rates. The simplest way to interact with the tool is to use the [Shiny App](https://portal.who.int/eios-colab/rconnect/simex).

The transmission model is written in the [**odin DSL**](https://mrc-ide.github.io/odin2/) in [`inst/odin/simex.R`](inst/odin/simex.R). That specification is compiled with [**dust2**](https://mrc-ide.github.io/dust2/) to C++ for fast discrete-time simulation (see the generated sources under `src/`). The main R entry point that uses this backend is `run_simex()`.

```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.cap = "",
  fig.align = "center",
  fig.path = "man/figures/",
  dpi = 200,
  out.width = "75%",
  message = FALSE,
  warning = FALSE,
  comment = ""
)
```

```{r, include = FALSE}
if (requireNamespace("pkgload", quietly = TRUE) && file.exists("DESCRIPTION")) {
  suppressPackageStartupMessages(pkgload::load_all("."))
} else {
  suppressPackageStartupMessages(library(simex))
}
suppressPackageStartupMessages(library(data.table))
```

Installation
-------------

To install the development version from GitHub:

```{r, eval = FALSE}
remotes::install_github("WHO-Collaboratory/simex")
```

Load the package using:

``` r
library(simex)
```

Running *simex*
-------------

### Shiny App

You can use *simex* interactively by launching the Shiny app locally with `simex::run_shiny()`. The app runs the same odin/dust2 model via `run_simex()`. If you want to use it programmatically, follow the steps below.

### Parameters and settings

Parameters are specified via `get_parameters()`. The arguments and their
default values are summarised below:

```{r, echo = FALSE, message = FALSE, warning = FALSE}
simex:::get_arg_table("get_parameters", type = "kable")
```

### Passing parameters to `run_simex()`

`run_simex()` accepts `pars` in three equivalent shapes. In each case you are
simulating **one** underlying system; only the layout of inputs differs.

1. **Single list from `get_parameters()`** — the same parameter values apply for
   the whole simulation (internally coerced to a one-column matrix keyed by
   `min(time)` from the `time` argument).

2. **Named list of parameter lists** — names must be **numeric strings** giving
   the **first model day** each parameter set applies from (e.g. `"1"`, `"75"`).
   One name must match the start of the simulated period (usually `"1"`).

3. **Matrix of lists** — **columns** are the same time breaks as in (2); **rows**
   are **parallel runs** (e.g. posterior draws). Outputs include a **`sample`**
   column indexing the matrix row; with `n_particles > 1`, a **`particle`**
   column indexes stochastic replicates.

For **different scenarios**, call **`run_simex()` separately** for each,
collect `simex` objects in a list, and pass it to **`vis_comparison()`** (e.g.
`lapply(list_of_par_lists, run_simex, time = 1:200)` or
`purrr::map(list_of_par_lists, \(p) run_simex(p, time = 1:200))`).

### Running default settings

Pass the list from `get_parameters()` to `run_simex()` and set `time` to the
integer days you want (inclusive).

```{r, echo = TRUE}
# set parameters using defaults
pars <- get_parameters()

# run odin/dust2 model (example: days 1–200)
output <- run_simex(pars, time = 1:200)

# look at output
print(output)
```

**`print()`** summarises **Time**, **Age categories**, **Compartments**
(`S`–`D`), **Particles** (default `1`), and **Samples** (default `1` when
`pars` has one row). See **Accessing outputs** for table columns.

### Visualising outputs

To visualise the results, use the generic `plot()` method for the `simex` class.
Below we visualise prevalence and then incidence via the `what` argument.

```{r, echo = TRUE}
# visualise prevalence
plot(output, what = "prevalence")

# visualise incidence
plot(output, what = "incidence")
```

Hospital capacity (when `what = "prevalence"`):

```{r, echo = TRUE}
plot(output, what = "prevalence", show_hosp_capacity = TRUE)
```

### Accessing outputs

Objects of class `simex` contain **`prevalence`** and **`incidence`**: long-format
`data.table`s with these columns:

* **`time`** — model day.
* **`age`** — age stratum.
* **`compartment`** — model compartment: `S`, `E`, `C`, `H`, `R`, or `D`.
* **`vax`** — logical indicating vaccination status.
* **`value`** — number of individuals.
* **`sample`** — present when `pars` has multiple matrix rows; indexes the row
  of `pars`.
* **`particle`** — present when `n_particles > 1`; indexes stochastic
  replicates.

The **`pars`** object passed into `run_simex()` is stored as **`output$pars`**.

Subset and aggregate with **`data.table`** syntax:

```{r, echo = TRUE}
# one day, one compartment, unvaccinated susceptible only
output$prevalence[time == 150L & compartment == "S" & !vax]
```

```{r, echo = TRUE}
# all strata between days 10 and 20 (inclusive)
output$prevalence[time %between% c(10L, 20L)]
```

```{r, echo = TRUE}
# total in hospital by day and vaccination status
output$prevalence[
  compartment == "H",
  .(hospitalised = sum(value)),
  by = .(time, vax)
]
```

The **`extract()`** helper remains available (e.g. credible intervals when
`sample` or `particle` are present); for simple shaping, prefer **`data.table`**
as above.

### Modelling a single intervention

Vaccination is controlled with **`vax_rate`** (doses per day, not a proportion of
the population).

```{r, echo = TRUE}
# vaccination rate: doses per day
pars <- get_parameters(vax_rate = 100)

output <- run_simex(pars, time = 1:200)

plot(output, what = "prevalence")
```

### Timed interventions (single system, parameters change over time)

One epidemic with isolation adherence stepping to `0.5` from day 75, using a
**named list** of parameter lists (equivalent to a **one-row matrix** with
column names `c("1", "75")`).

```{r, echo = TRUE}
pars_timed <- list(
  "1" = get_parameters(),
  "75" = get_parameters(isolation_adherence = 0.5)
)

output_timed <- run_simex(pars_timed, time = 1:200)

plot(output_timed, what = "prevalence")
```

### Comparing different scenarios (separate runs)

Two **different** scenarios require two **`run_simex()`** calls, not a
multi-row `pars` matrix. Pass a **named list** of `simex` objects to
**`vis_comparison()`**.

```{r, echo = TRUE}
scenario_pars <- list(
  "No intervention" = get_parameters(),
  "Isolation from day 75" = list(
    "1" = get_parameters(),
    "75" = get_parameters(isolation_adherence = 0.5)
  )
)

outputs <- lapply(scenario_pars, run_simex, time = 1:200)

vis_comparison(outputs)
```

The same pattern works with **`purrr::map()`**, e.g.
`purrr::map(scenario_pars, \(p) run_simex(p, time = 1:200))`.

## Contributors

- [Finlay Campbell](https://github.com/finlaycampbell) (campbellf@who.int)
- Prabasaj Paul (ppaul@who.int)

**Maintainer:** Finlay Campbell

## Licensing

The simex software is made available by Collaboratory (2024) under an [MIT license](LICENSE.md) ([citation file](CITATION.cff)).

The demographic data used in this software is made available by United Nations (2024) under a [Creative Commons license CC BY 3.0 IGO](http://creativecommons.org/licenses/by/3.0/igo/). Source:
> United Nations, Department of Economic and Social Affairs, Population Division (2024). [World Population Prospects 2024, Online Edition](https://population.un.org/wpp/).

The contact data used in this software is made available by Prem et al. (2017) under a [Creative Commons license CC BY 4.0](https://creativecommons.org/licenses/by/4.0/). Source:
> Kiesha Prem, Alex R. Cook, Mark Jit, *Projecting social contact matrices in 152 countries using contact surveys and demographic data*, PLoS Comp. Biol. (2017), https://doi.org/10.1371/journal.pcbi.1005697.
