*simex*: a disease modelling tool for decision makers
---------------------------------------------------------------

[![MIT license](https://img.shields.io/badge/License-MIT-blue.svg)](LICENSE.md) [![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://lifecycle.r-lib.org/articles/stages.html#maturing)

*simex* is an R package for simulating the spread of infectious diseases, as well as interventions such as social distancing measures, isolation and vaccination. It uses an age-structured SEIR compartmental model with country-specific age demographics and contact rates. The simplest way to interact with the tool is to use the [Shiny App](https://portal.who.int/eios-colab/rconnect/simex).

The transmission model is written in [**odin**](https://mrc-ide.github.io/odin/) (odin2 DSL) in [`inst/odin/simex.R`](inst/odin/simex.R). That specification is compiled with [**dust2**](https://mrc-ide.github.io/dust2/) to C++ for fast discrete-time simulation (see the generated sources under `src/`). The main R entry point that uses this backend is `run_simex()`.

```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.cap = "",
  fig.align = "center",
  fig.path = "man/figures/",
  dpi = 200,
  out.width = "75%",
  message = FALSE,
  warning = FALSE,
  comment = ""
)
```

```{r, include = FALSE}
if (requireNamespace("pkgload", quietly = TRUE) && file.exists("DESCRIPTION")) {
  suppressPackageStartupMessages(pkgload::load_all("."))
} else {
  suppressPackageStartupMessages(library(simex))
}
suppressPackageStartupMessages(library(data.table))
```

Installation
-------------

To install the development version from github:

```{r, eval = FALSE}
remotes::install_github("WHO-Collaboratory/simex")
```

Load the package using:

``` r
library(simex)
```

Running *simex*
-------------

### Shiny App
You can use *simex* interactively by launching the Shiny app locally with `simex::run_shiny()`. The app runs the same odin/dust2 model via `run_simex()`. If you want to use it programmatically, follow the steps below.

### Parameters and settings

Most settings are specified via `get_parameters()`. The arguments and their
default values are summarised below (from the roxygen documentation):

```{r, echo = FALSE, message = FALSE, warning = FALSE}
simex:::get_arg_table("get_parameters", type = "kable")
```

### Passing parameters to `run_simex()`

`run_simex()` accepts `pars` in three equivalent shapes. In each case you are
simulating **one** underlying system (one epidemic process); only the layout
of inputs differs.

1. **Single list from `get_parameters()`** — the same parameter values apply for
   the whole simulation. Internally this is turned into a one-column matrix
   whose column name is the first time in `time`.

2. **Named list of parameter lists** — names must be **numeric strings** giving
   the **first model day** each parameter set applies from (e.g. `"1"`,
   `"75"`). The list is coerced to a one-row matrix with those names as column
   names. One name must correspond to the start of the simulation (typically
   `"1"`).

3. **Matrix of lists** — **columns** are the same time breaks as in (2): each
   column is one parameter list for that segment. **Rows** are **parallel
   runs** of the same calendar schedule (e.g. rows = posterior draws). The
   simulator uses `dust2` groups so each row is independent; outputs then
   include a **`sample`** column indexing the row (parallel draw). With
   `n_particles > 1` you also get a **`particle`** column for stochastic
   replicates within each run.

To compare **different scenarios as different models** (e.g. “no vaccine” vs
“high vaccine”), run **`run_simex()` separately** for each scenario and collect
the `simex` objects in a list, then pass that list to `vis_comparison()`. For
example: `lapply(list_of_par_lists, run_simex, time = 1:200)` or, with
**`purrr`**, `purrr::map(list_of_par_lists, \(p) run_simex(p, time = 1:200))`.

### Running default settings

Pass the list from `get_parameters()` to `run_simex()` and set `time` to the
integer days you want (inclusive).

```{r, echo = TRUE}
# set parameters using defaults
pars <- get_parameters()

# run odin/dust2 model (example: days 1–200)
output <- run_simex(pars, time = 1:200)

# look at output
print(output)
```

The **`print()`** summary lines mean: **Time** — range of simulated days;
**Age categories** — factor levels in the output tables; **Compartments**
— epidemic stages (`S` … `D`); **Particles** — stochastic replicates
(`1` with the default single particle; larger when `n_particles > 1`);
**Samples** — parallel parameter groups (`1` with a single-row `pars`;
larger when `pars` has multiple matrix rows, e.g. posterior draws).

In **`output$prevalence`** and **`output$incidence`** (both `data.table`s):

* **`time`** — model day.
* **`age`** — age stratum (`age_1`, …).
* **`compartment`** — `S`, `E`, `C`, `H`, `R`, or `D`.
* **`vax`** — `TRUE` / `FALSE` for vaccinated vs unvaccinated strata.
* **`value`** — count in that stratum (or flow for incidence).
* **`sample`** — present when `pars` is a matrix with **multiple rows**
  (parallel draws); indexes the row of `pars`.
* **`particle`** — present when `n_particles > 1`.

### Visualising outputs

To visualise the results, use the generic `plot` function defined for the `simex` class. Below, we first visualise prevalence and then incidence, specified using the `what` argument.

```{r, echo = TRUE}
# visualise prevalence
plot(output, what = "prevalence")

# visualise incidence
plot(output, what = "incidence")
```

Hospital capacity can be displayed by toggling the `show_hosp_capacity` argument.

```{r, echo = TRUE}
# visualise prevalence with hospital capaciy
plot(output, what = "prevalence", show_hosp_capacity = TRUE)
```

### Accessing outputs

Objects returned by `run_simex()` have class `simex` and contain:

* **`prevalence`** — long-format `data.table` (see column definitions above).
* **`incidence`** — same layout; **`value`** is the daily flow into each
  compartment (see `?extract` for interpretation by letter).
* **`pars`** — the `pars` object passed into `run_simex()` (often a matrix of
  lists when using time-varying or multi-row input).

Subset and aggregate with **`data.table`** syntax on `output$prevalence` or
`output$incidence`:

```{r, echo = TRUE}
# one day, one compartment, unvaccinated susceptible only
output$prevalence[time == 150L & compartment == "S" & !vax]
```

```{r, echo = TRUE}
# all strata between days 10 and 20 (inclusive)
output$prevalence[time %between% c(10L, 20L)]
```

```{r, echo = TRUE}
# total in hospital by day and vaccination status
output$prevalence[
  compartment == "H",
  .(hospitalised = sum(value)),
  by = .(time, vax)
]
```

The **`extract()`** helper is still available for convenience (e.g. credible
intervals when `sample` or `particle` are present); for simple shaping,
prefer **`data.table`** on the stored tables as above.


### Modelling a single intervention

We consider vaccination as an example intervention, using the
`vax_rate` argument to specify the number of doses per day.

```{r, echo = TRUE}
# define vaccination rate (doses per day)
pars <- get_parameters(vax_rate = 100)

# run model
output <- run_simex(pars, time = 1:200)

# visualise prevalence
plot(output)
```

We can see that the daily increase in number of vaccinated individuals, as
well as the impact on infection and disease severity.

### Timed interventions (single system, parameters change over time)

Below, **one** epidemic is simulated with isolation adherence stepping to `0.5`
from day 75. This uses a **named list** of parameter lists; a **one-row matrix**
with column names `c("1", "75")` would be equivalent.

```{r, echo = TRUE}
pars_timed <- list(
  "1" = get_parameters(),
  "75" = get_parameters(isolation_adherence = 0.5)
)

output_timed <- run_simex(pars_timed, time = 1:200)

plot(output_timed, what = "prevalence")
```

### Comparing different scenarios (separate runs)

Here we simulate **two different scenarios** as **two calls** to `run_simex()`
and collect two `simex` objects. That differs from a multi-column `pars`
matrix, which varies parameters over time for **one** scenario with parallel
draws across rows.

`vis_comparison()` takes a **named list** of `simex` objects (one per scenario).

```{r, echo = TRUE}
scenario_pars <- list(
  "No intervention" = get_parameters(),
  "Isolation from day 75" = list(
    "1" = get_parameters(),
    "75" = get_parameters(isolation_adherence = 0.5)
  )
)

outputs <- lapply(scenario_pars, run_simex, time = 1:200)

vis_comparison(outputs)
```

The same idea works with **`purrr::map()`** if you prefer a tidyverse style,
e.g. `purrr::map(scenario_pars, \(p) run_simex(p, time = 1:200))`.

### Contributors
------------
- [Finlay Campbell](https://github.com/finlaycampbell) (campbellf@who.int)
- Prabasaj Paul (ppaul@who.int)

**Maintainer:** Finlay Campbell


### Licensing

The simex software is made available by Collaboratory (2024) under an [MIT license](LICENSE.md) ([citation file](CITATION.cff)).

The demographic data used in this software is made available by United Nations (2024) under a [Creative Commons license CC BY 3.0 IGO](http://creativecommons.org/licenses/by/3.0/igo/). Source:
> United Nations, Department of Economic and Social Affairs, Population Division (2024). [World Population Prospects 2024, Online Edition](https://population.un.org/wpp/).

The contact data used in this software is made available by Prem et al. (2017) under a [Creative Commons license CC BY 4.0](https://creativecommons.org/licenses/by/4.0/). Source:
> Kiesha Prem, Alex R. Cook, Mark Jit, *Projecting social contact matrices in 152 countries using contact surveys and demographic data*, PLoS Comp. Biol. (2017), https://doi.org/10.1371/journal.pcbi.1005697.
