*simex*: a disease modelling tool for decision makers
---------------------------------------------------------------

[![MIT license](https://img.shields.io/badge/License-MIT-blue.svg)](https://github.com/epiforecasts/EpiNow2/blob/main/LICENSE.md/) [![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://lifecycle.r-lib.org/articles/stages.html#maturing)

*simex* is an R package for simulating the spread of infectious diseases, as
well as interventions such as social distancing measures, isolation and
vaccination. It uses an age-structured SEIR compartmental model with
country-specific age demographics and contact rates.

```{r, echo = FALSE} 
knitr::opts_chunk$set(fig.cap="",
                      fig.align = "center",
                      dpi = 200,
                      out.width = "75%")
``` 

Installation
-------------

To install the development version from github:

```{r, eval = FALSE}
remotes::install_github("finlaycampbell/simex", dependencies = TRUE, force = TRUE)
```

Load the package using:

```{r}
library("simex")
```

Running *simex*
-------------

### Parameters and settings

Most settings are specified via the `get_parameters` function. The arguments and
their default values are given below:

```{r, echo = FALSE}
simex:::get_arg_table("get_parameters", type = "kable")
```

### Running default settings

To run the model using default settings, specify a parameter object `par` and
feed this into the `run_model` function.

```{r, echo = TRUE}
## set parameters using defaults
pars <- get_parameters()

## run model using default parameters
output <- run_model(pars)

## look at output
print(output)
```

### Visualising outputs

To visualise the results, use the generic `plot` function defined for the
`simex` class. Below, we first visualise prevalence and then incidence,
specified using the `what` argument.

```{r, echo = TRUE}
## visualise prevalence
plot(output, what = "prevalence")

## visualise incidence
plot(output, what = "incidence")
```

Hospital capacity can be displayed by toggling the `show_hosp_capacity`
argument.

```{r, echo = TRUE}
## visualise prevalence with hospital capaciy
plot(output, what = "prevalence", show_hosp_capacity = TRUE)
```

### Summarising outputs

To get a summary of the model state at a given point in time, use the `summary`
function with the optional `day` argument. In the example below, we display the
summary statistics for day 150:

```{r, eval = FALSE}
## get summary at day 150
summary(output, day = 150)
```
```{r, echo = FALSE}
## kable for Rmd
knitr::kable(summary(output, day = 150))
```



### Accessing outputs

The `output` object is of class `simex` and is a list containing four items:

* `prevalence` is an array with 3 dimensions: time (365 days) x age (16
  categories) x state (12 compartments). The compartments are S (susceptible), E
  (exposed, pre-symptomatic), C (symptomatic in community), H (sympomatic in
  hospital), R (recovered), D (dead). Each comparment is split into unvaccinated
  (with subscript _u) and vaccinated (with subscript _v). Each value in the
  array represents the proportion of the population in that given age-disease
  compartment on a given day.
* `deltas` is an array with the same dimensions as `prevalence`. Each value
  represents the change in a given age-disease compartment from the previous
  day.
* `incidence` is an array with the same dimensions as `prevalence`. Each value
  represents the new additions to a given age-disease comparment on a given
  day. For the exposed category E, this represents incidence of new infections,
  for the hospitalised category H this represents new hospital admissions, and
  for the dead category D this represents new deaths.
* `pars` contains the parameter set initially fed into `run_model`.

Accessing the data is easiest using array indexing. Remember, the dimensions are
time, age and compartment. For example, accessing the prevalence on the 250th
day is done with `output$prevalence[250,,]`:

```{r, echo = FALSE}
## extract prevalence in percent on day 250
round(100*output$prevalence[250,,], 1)
```

Accessing the prevalence of the 1st age compartment (0-4) and 1st infectious
compartment (unvaccinated susceptible) for days 130 to 135 is done with
`output$prevalence[130:135,1,1]`:

```{r, echo = FALSE}
## extract in percent on day 130-135
round(100*output$prevalence[130:135,1,1], 1)
```

The outputs can also be extracted in `tibble` form using the `extract` function,
once again using the `what` argument to specify whether prevalence or incidence
is extracted.

```{r, echo = TRUE}
## extract prevalence
extract(output, what = "prevalence")
```
If we want to filter this list for a sequence of days, we can then do basic
dataframe manipulation:

```{r, echo = TRUE}
## define start and end days
days_from <- 10
days_to <- 20

## extract prevalence
df <- extract(output, what = "prevalence")
df <- df[df$day %in% seq(days_from, days_to),]
df
```

### Modelling a single intervention

We use vaccination as an example intervention. Referencing the table above, we
can see that vaccination rate is specified using the `vax_rate` argument and
set it to 0.5% of the population per day.

```{r, echo = TRUE}
## define vaccination rate
pars <- get_parameters(vax_rate = 0.005)

## run model
output <- run_model(pars)

## visualise prevalence
plot(output)
```

We can see that the daily increase in number of vaccinated individuals, as
well as the impact on infection and disease severity.


### Specifying an initial state

The default initial state begins with an entirely susceptible, unvaccinated
population and a single infection. We can also specify a custom initial state by
passing a matrix specifying the number of individuals in each age-infection
compartment. In the example below, we extract the initial state from the
previous run, and modify the compartments so half the susceptible population is
assigned to the vaccinated compartment.

```{r, echo = TRUE}
## extract starting point from previous run
state <- output$prevalence[1,,]

## assign half of susceptibles to vaccinated
state[,"S_u"] <- state[,"S_v"] <- state[,"S_u"]/2

## run model
output <- run_model(pars, init_state = state)

## visualise prevalence
plot(output)
```

We can see that the simulation now begins with a 50% vaccinated population,
significantly reducing the size and impact of the pandemic.

### Timed interventions, multiple interventions

Sometime we only want to introduce interventions at a certain time, or we want
to model multiple, successive interventions. We can do this easily by generating
a named list of parameters, where the name gives the time that parameter set
should be used from. In the example below, we introduce isolation measures with
an adherence of 50% on the 125th day.

```{r, echo = TRUE}
## introduce isolation on the 125th day
parlist <- list(
  "1" = get_parameters(),
  "125" = get_parameters(isolation_adherence = 0.5)
)

## run model
output <- run_model(parlist)

## visualise prevalence
plot(output)
```

Comparing this figure with the first model run with no interventions, we can see
the proportion of deaths drops from about 0.7% to 0.4%; a reduction in deaths of
more than 40%!

### Comparing scenarios

It is useful to directly compare different scenarios visually. The
`vis_comarison` function does exactly this, and accepts a named list of `simex`
objects. In the example below, we generate a named list of lists that compares
the default scenario (no interventions) with the a scenario where isolation is
introduced on the 125th day.
	
```{r, echo = TRUE}
## define two scenarios, one without intervention and one with isolation
parlists <- list(
  "No intervention" = get_parameters(),
  "Isolation on day 125" = list(
    "1" = get_parameters(),
    "125" = get_parameters(isolation_adherence = 0.5)
  )
)

## run model across set of parameter lists
outputs <- lapply(parlists, run_model)

## compare scenarios
vis_comparison(outputs)
```

### Contributors
------------
- [Finlay Campbell](https://github.com/finlaycampbell) (campbellf@who.int)
- Prabasaj Paul (ppaul@who.int)

**Maintainer:** Finlay Campbell


### Licensing

The simex software is made available under an [MIT license](LICENSE.md). [Suggested citation](CITATION.cff).

The demographic data used in this software is made available by United Nations (2024) under a [Creative Commons license CC BY 3.0 IGO](http://creativecommons.org/licenses/by/3.0/igo/).
> United Nations, Department of Economic and Social Affairs, Population Division (2024). [World Population Prospects 2024, Online Edition](https://population.un.org/wpp/).

The contact data used in this software is made available by Prem et al. (2017) under a [Creative Commons license CC BY 4.0](https://creativecommons.org/licenses/by/4.0/).
> Kiesha Prem, Alex R. Cook, Mark Jit, *Projecting social contact matrices in 152 countries using contact surveys and demographic data*, PLoS Comp. Biol. (2017), https://doi.org/10.1371/journal.pcbi.1005697.
