*simex*: a disease modelling tool for decision makers
---------------------------------------------------------------

[![MIT license](https://img.shields.io/badge/License-MIT-blue.svg)](LICENSE.md) [![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://lifecycle.r-lib.org/articles/stages.html#maturing)

*simex* is an R package for simulating the spread of infectious diseases, as well as interventions such as social distancing measures, isolation and vaccination. It uses an age-structured SEIR compartmental model with country-specific age demographics and contact rates. The simplest way to interact with the tool is to use the [Shiny App](https://portal.who.int/eios-colab/rconnect/simex).

The transmission model is written in [**odin**](https://mrc-ide.github.io/odin/) (odin2 DSL) in [`inst/odin/simex.R`](inst/odin/simex.R). That specification is compiled with [**dust2**](https://mrc-ide.github.io/dust2/) to C++ for fast discrete-time simulation (see the generated sources under `src/`). The main R entry point that uses this backend is `run_simex()`.

```{r, echo = FALSE} 
knitr::opts_chunk$set(fig.cap = "",
                      fig.align = "center",
                      fig.path = "man/figures/",
                      dpi = 200,
                      out.width = "75%")
``` 

Installation
-------------

To install the development version from github:

```{r, eval = FALSE}
remotes::install_github("WHO-Collaboratory/simex")
```

Load the package using:

```{r}
library("simex")
```

Running *simex*
-------------

### Shiny App
You can use *simex* interactively by launching the Shiny app locally with `simex::run_shiny()`. The app runs the same odin/dust2 model via `run_simex()`. If you want to use it programmatically, follow the steps below.

### Parameters and settings

Most settings are specified via `get_parameters()`. The arguments and their
default values are summarised below (from the roxygen documentation):

```{r, echo = FALSE}
simex:::get_arg_table("get_parameters", type = "kable")
```

### Running default settings

To run the model using default settings, pass the list from `get_parameters()` to
`run_simex()` and set `time` to the integer days you want (inclusive).

```{r, echo = TRUE}
## set parameters using defaults
pars <- get_parameters()

## run odin/dust2 model (example: days 1–365)
output <- run_simex(pars, time = 1:365)

## look at output
print(output)
```

### Visualising outputs

To visualise the results, use the generic `plot` function defined for the `simex` class. Below, we first visualise prevalence and then incidence, specified using the `what` argument.

```{r, echo = TRUE}
## visualise prevalence
plot(output, what = "prevalence")

## visualise incidence
plot(output, what = "incidence")
```

Hospital capacity can be displayed by toggling the `show_hosp_capacity` argument.

```{r, echo = TRUE}
## visualise prevalence with hospital capaciy
plot(output, what = "prevalence", show_hosp_capacity = TRUE)
```

### Accessing outputs

Objects returned by `run_simex()` have class `simex` and are a list with at
least:

* `prevalence` — long-format `data.table` with columns including `time`, `age`,
  `compartment` (`S`, `E`, `C`, `H`, `R`, `D`), `vax`, and `value`. Stochastic
  runs add `particle` and/or `sample` when relevant.
* `incidence` — same layout for daily flows into each stage (see `?extract`).
* `pars` — parameter object(s) passed into `run_simex()`.

Example: prevalence on day 250 for unvaccinated susceptibles:

```{r, echo = TRUE}
subset(output$prevalence, time == 250 & compartment == "S" & vax == FALSE)
```

Use `extract()` to aggregate over dimensions:

```{r, echo = TRUE}
## long-form prevalence (default stratification)
extract(output, what = "prevalence")
```

```{r, echo = TRUE}
## restrict to a band of days
days_from <- 10
days_to <- 20
df <- extract(output, what = "prevalence")
df <- df[df$time %in% seq(days_from, days_to), ]
df
```

### Modelling a single intervention

We use vaccination as an example intervention. Referencing the table above, we
can see that vaccination rate is specified using the `vax_rate` argument and
set it to 0.5% of the population per day.

```{r, echo = TRUE}
## define vaccination rate
pars <- get_parameters(vax_rate = 0.005)

## run model
output <- run_simex(pars, time = 1:365)

## visualise prevalence
plot(output)
```

We can see that the daily increase in number of vaccinated individuals, as
well as the impact on infection and disease severity.

### Timed interventions, multiple interventions

To change parameters part-way through a run, pass a one-row **matrix** of
parameter lists whose **column names** are the first model day each block applies
to (one column must start on day 1, matching the Shiny app). Below, isolation
adherence becomes 0.5 from day 125 onward.

```{r, echo = TRUE}
## column names = first day each parameter set applies
parlist <- matrix(
  c(list(get_parameters()), list(get_parameters(isolation_adherence = 0.5))),
  nrow = 1,
  dimnames = list(NULL, c("1", "125"))
)

output <- run_simex(parlist, time = 1:365)

## visualise prevalence
plot(output)
```

Comparing this figure with the first model run with no interventions, we can see
the proportion of deaths drops from about 0.7% to 0.4%; a reduction in deaths of
more than 40%!

### Comparing scenarios

It is useful to directly compare different scenarios visually. The
`vis_comparison()` function accepts a named list of `simex` objects. In the
example below, the default scenario is compared with isolation from day 125.

```{r, echo = TRUE}
## define two scenarios, one without intervention and one with isolation
parlists <- list(
  "No intervention" = get_parameters(),
  "Isolation on day 125" = matrix(
    c(list(get_parameters()), list(get_parameters(isolation_adherence = 0.5))),
    nrow = 1,
    dimnames = list(NULL, c("1", "125"))
  )
)

outputs <- lapply(parlists, run_simex, time = 1:365)

## compare scenarios
vis_comparison(outputs)
```

### Contributors
------------
- [Finlay Campbell](https://github.com/finlaycampbell) (campbellf@who.int)
- Prabasaj Paul (ppaul@who.int)

**Maintainer:** Finlay Campbell


### Licensing

The simex software is made available by Collaboratory (2024) under an [MIT license](LICENSE.md) ([citation file](CITATION.cff)).

The demographic data used in this software is made available by United Nations (2024) under a [Creative Commons license CC BY 3.0 IGO](http://creativecommons.org/licenses/by/3.0/igo/). Source:
> United Nations, Department of Economic and Social Affairs, Population Division (2024). [World Population Prospects 2024, Online Edition](https://population.un.org/wpp/).

The contact data used in this software is made available by Prem et al. (2017) under a [Creative Commons license CC BY 4.0](https://creativecommons.org/licenses/by/4.0/). Source:
> Kiesha Prem, Alex R. Cook, Mark Jit, *Projecting social contact matrices in 152 countries using contact surveys and demographic data*, PLoS Comp. Biol. (2017), https://doi.org/10.1371/journal.pcbi.1005697.
